---
title: "Reproducible Research Course Project 1"
author: "Olga Dergachyova"
date: "November 2018"
output:
    md_document:
        variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale("LC_TIME", "en_US.UTF-8")
library(lattice)
```

## Loading and preprocessing the data

```{r loading}
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
Note: the missing values in the dataset are ignored.

1. Calculate the total number of steps taken per day

```{r total steps per day}
data_by_day <- aggregate(steps ~ date, data, sum)
```

2. Make a histogram of the total number of steps taken each day

```{r histogram}
hist(data_by_day$steps, main = paste("Total number of steps per day"), xlab="Number of steps", ylab="Number of days", col="green")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r mean and median}
steps_mean <- round(mean(data_by_day$steps), 2)
steps_median <- median(data_by_day$steps)
```

The mean total number of steps per day is `r toString(steps_mean)`, and median is `r toString(steps_median)`.

## What is the average daily activity pattern?

1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time series plot}
data_by_interval <- aggregate(steps ~ interval, data, mean)
plot(data_by_interval$interval, data_by_interval$steps, type="l", xlab="Interval", ylab="Average number of steps", main="Average number of steps per day and per interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max steps}
maxstep_interval <- data_by_interval$interval[which.max(data_by_interval$steps)]
```

The interval containing the maximum number of steps on average is `r maxstep_interval`.

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)

```{r missing values}
total_na <- sum(is.na(data))
total_na
```

2. Create a new dataset by filling in missing values in the current dataset with means for corresponding 5-minute intervals. 

```{r missing intervals}
filled_data <- data.frame(data)
missing_values = which(is.na(filled_data$steps))
filled_data[missing_values, "steps"] <- sapply(filled_data[missing_values, "interval"], function(x){
    x <- data_by_interval$steps[data_by_interval$interval==x]
})
```

3. Make a histogram of the total number of steps taken each day for new filled in data

```{r histogram filled data}
filled_data_by_day <- aggregate(steps ~ date, filled_data, sum)
hist(filled_data_by_day$steps, main = paste("Total number of steps per day in filled in data"), xlab="Number of steps", ylab="Number of days", col="blue")
```

4. Calculate and report the mean and median of the total number of steps taken per day for filled in data

```{r mean and median filled data}
steps_mean <- round(mean(filled_data_by_day$steps), 2)
steps_median <- round(median(filled_data_by_day$steps), 2)
```
In filled data the mean total number of steps per day is `r toString(steps_mean)`, and median is `r toString(steps_median)`. With the addressed filling in strategy, the median value is now equal to mean value.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r weekday weekend}
filled_data$day <- ifelse(weekdays(as.Date(filled_data$date)) %in% c("Saterday", "Sunday"), 1, 0)
filled_data$day <- factor(filled_data$day, labels = c("weekday", "weekend"))
```

2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r panel plot}
data_interval_days <- aggregate(steps ~ interval + day, filled_data, mean)
xyplot(data_interval_days$steps ~ data_interval_days$interval|data_interval_days$day, type="l", main="Average number of steps per day type and per interval", xlab="Interval", ylab="Average number of steps", layout=c(1,2))
```
